---
title: "EmmaBeyer_A02_CodingBasics.Rmd"
author: "Emma Beyer"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq(1, 30, 3) #sequence from 1 to 30 increasing by 3
A02_sequence_1to30 <- seq(1, 30, 3) #assigning name to sequence

#2. 
mean(A02_sequence_1to30) #mean of sequence
median(A02_sequence_1to30) #median of sequence

#3. 
mean(A02_sequence_1to30) > median(A02_sequence_1to30) #determines if mean is greater than median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
names_of_students <- c("Harry", "Ron", "Hermione", "Neville", "Luna", "Draco") #vector of student names
test_scores <- c(70, 50, 100, 60, 80, 40) #vector of test scores from students
passing_grade <- test_scores > 50 #determines if test scores are above 50

student_grades_df <- data.frame("Student_Names"=names_of_students, "Test_Scores"=test_scores, "Passing_Grade"=passing_grade) #
student_grades_df #creating a data frame with students, grades, and their passing grades

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:A matrix must have the same modes and the same length for all columns, but a data frame can have different modes.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
