---
title: "EmmaBeyer_A05_DataVisualization.Rmd"
author: "Emma Beyer"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
# loading packages
library(tidyverse); library(lubridate); library(here); library(cowplot); library(ggthemes)
# checking working directory
here()
# reading data
PeterPaul.nutrients <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
           stringsAsFactors = TRUE)
PeterPaul.litter <-
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
           stringsAsFactors = TRUE)                                              #!!!!!!

#2 
# check dates
class(PeterPaul.nutrients$sampledate)
class(PeterPaul.litter$collectDate)
# change factors to dates
PeterPaul.nutrients$sampledate <- ymd(PeterPaul.nutrients$sampledate)
PeterPaul.litter$collectDate <- ymd(PeterPaul.litter$collectDate)
# recheck dates
class(PeterPaul.nutrients$sampledate)
class(PeterPaul.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
# assigning default theme
my_theme <- theme_base()  + 
        # changing background color
  theme(panel.background =  element_rect(color = 'lightgrey'),
        # changing legend position
        legend.position = 'bottom right',
        # changing axis lines to blue
        panel.grid.major = element_line(color = "blue"),
        panel.grid.minor = element_line(color = "blue"))

# setting my_theme as default
theme_set(my_theme)
# ask if I need to add this at the end of my code or if auto                    !!!!!!
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
tp_po4_plot <- ggplot(filter(PeterPaul.nutrients),
              # plotting phosphorus by phosphate and plotting lakes with different colors
              aes(y=tp_ug,
                  x=po4,
                  color=lakename)) +
              # adding line of best fit colored black
              geom_smooth(method = "lm",
                          color="black") +
              # changing the limits
              xlim(0, 250) +
              ylim(0, 20) +
              # adding default theme
              my_theme()
print(tp_po4_plot)
# need to change the limits once my data is correct                             !!!!!!
# need to check to make sure mytheme() works once I figure that out             !!!!!!
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
temp_plot <- ggplot(filter(PeterPaul.nutrients),
           # plotting temperature by month
            aes(y=temperature_C,
                x=factor(month,
                         levels=1:12,
                         labels=month.abb))) +
          # choosing box plot
          geom_boxplot() +
          # adding x axis name
          scale_x_discrete(name="month",
                           drop=FALSE)

tp_plot <- ggplot(filter(PeterPaul.nutrients),
           # plotting temperature by month
            aes(y=tp_ug,
                x=factor(month,
                         levels=1:12,
                         labels=month.abb))) + 
          # choosing box plot
          geom_boxplot() +
          # adding x axis name
          scale_x_discrete(name="month",
                           drop=FALSE)

tn_plot <- ggplot(filter(PeterPaul.nutrients),
           # plotting temperature by month
            aes(y=tn_ug,
                x=factor(month,
                         levels=1:12,
                         labels=month.abb))) + 
          # choosing box plot
          geom_boxplot() +
          # adding x axis name
          scale_x_discrete(name="month",
                           drop=FALSE)

# adding above graphs to one plot
plot_grid(temp_plot, tp_plot, tn_plot, nrow = 2, align = 'h', rel_heights = c(1.25, 1))

  
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
# displaying only Needles subset
needles_plot <- ggplot(filter(PeterPaul.litter, functionalGroup=="Needles"),
                     # plotting dry mass of needles by date and coloring nlcdClass
                     aes(y=dryMass,
                         x=collectDate,
                         color=nlcdClass))
                        # choosing violin plot
                        geom_violin() +
                        # changing the name of the legend
                        labs(color = "NLCD Class")
print(needles_plot)
# ask is there a way to fill the violins                                        !!!!!!

#7
# displaying only Needles subset
needles_plot_facet <- ggplot(filter(PeterPaul.litter, functionalGroup=="Needles"),
                          # plotting dry mass of needles by date and coloring nlcdType
                          aes(y=dryMass,
                              x=collectDate,)) +
                          # choosing violin plot
                          geom_violin() +
                          # separate NLCD classes into 3 facets
                          facet_wrap(vars(nlcdClass), nrow = 3, )
print(needles_plot_facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: It's easier to compare the NLCD Classes in 6, since they are next to each other on the same plot. Separated the graph gets squished and it's harder to see the differences between them.