---
title: "EmmaBeyer_A07_GLM.Rmd"
author: "Emma Beyer"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
# set libraries and check WD
library(tidyverse)
library(agricolae)
library(corrplot)
library(dplyr)
library(here)
here()

# load data
ChemPhys <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

# Set date to date format, would give back NAs
#ChemPhys$sampledate <- as.Date(ChemPhys$sampledate, format = "%m-%d-%y")
# now shows Y,m,d
ChemPhys$sampledate <- mdy(ChemPhys$sampledate)

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "darkgreen"), 
        title = element_text(color='darkblue'),
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: pvalue >= 0.05
Ha: pvalue <= 0.05


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ChemPhys_subset <-
  ChemPhys %>%
  # only dates in July
  filter(month(sampledate) == 7) %>%
  # select columns
  select(lakename, year4, daynum, depth, temperature_C) %>%
  # omitting NAs
  na.omit()

#5
tempbydepth_scatter <- 
  ggplot(ChemPhys_subset, aes(x = depth, 
                              y = temperature_C)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = "red") +
  ylim(0,35) +
  labs(x = "Depth",
       y = "Temperature C",
       title = "Temperature vs. Depth in July")
print(tempbydepth_scatter)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It suggest that as depth increases the temperature decreases. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
# run linear regression
depth_temp_regression <- lm(
  data = ChemPhys_subset,
  temperature_C ~ depth)
summary(depth_temp_regression)
# p-value < 2.2e-16
# r-squared = 0.7387 

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: About 73.9% of the variablity in temperature is explained by the changes in depth. There are 9726 degrees of freedom. This relationship is significant because the pvalue is less than 0.05, which indicated a significant relationship between temperature and depth. The slope of the regression line is -1.95, which shows that the temperature is decreasing by 1.95 degrees every meter. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
# wrangling data
ChemPhys_AIC_subset <- ChemPhys %>%
  # only dates in July
  filter(month(sampledate) == 7) %>%
  # select columns
  select(year4, daynum, depth, temperature_C) %>%
  # omitting NAs
  na.omit()

# running AIC
ChemPhys_Corr <- cor(ChemPhys_AIC_subset)
# full plot
corrplot(ChemPhys_Corr, method = "ellipse")
# top half of plot
corrplot.mixed(ChemPhys_Corr, upper = "ellipse")

#10
# running multiple regression 
ChemPhys_AIC_BEST_regression <- lm(data = ChemPhys_AIC_subset, temperature_C ~ depth + year4 + daynum)
# summary of multiple regression
summary(ChemPhys_AIC_BEST_regression)

# scatter plot of multiple regression
multiregression_plot <- ggplot(ChemPhys_subset, 
                               aes(x = depth, 
                               y = temperature_C,
                               color = daynum)) +
  # setting scatter plot
  geom_point() +
  # setting y limit
  ylim(0, 35) +
  # adding labels
  labs(x = "Depth",
       y = "Temperature C",
       title = "Temperature vs. Depth in July by Daynum")
print(multiregression_plot)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: It suggests that we use depth and daynum to predict temperature. This explains 74.1% of the observed variance. Adding daynum improves the model by a little bit, only 0.2%. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# format as aov
ChemPhys_anova <- aov(data = ChemPhys_subset, temperature_C ~ lakename)
summary(ChemPhys_anova)

# format as lm
ChemPhys_anova2 <- lm(data = ChemPhys_subset, temperature_C ~ lakename)
summary(ChemPhys_anova2)

plot(ChemPhys_anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Each lake had a pvalue less than 0.05, so they was a significant difference in mean temperature at all lakes. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
# graph of temp by depth
templakename_scatter <- 
  # setting variables
  ggplot(ChemPhys_subset, aes(x = depth, 
                              y = temperature_C,
                              color = lakename)) +
  # creating scatter plot
  geom_point() +
  # adding regression line
  geom_smooth(method = 'lm', se = FALSE, aes(group = lakename)) +
  # setting y limits
  ylim(0,35) +
  # adding labels
  labs(x = "Depth",
       y = "Temperature C",
       title = "Temperature vs. Depth in July across Lakes")
print(templakename_scatter)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# running Tukey's HSD test
TukeyHSD(ChemPhys_anova)


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# wrangling data
ChemPhys_lake_subset <- ChemPhys %>%
  # only dates in July
  filter(month(sampledate) == 7) %>%
  # only include Crampton and Ward Lake
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake") %>%
  # select columns
  select(year4, daynum, depth, temperature_C, lakename) %>%
  # omitting NAs
  na.omit()
print(ChemPhys_lake_subset)

# two-sample T-test
lake_twosample <- t.test(ChemPhys_lake_subset$temperature_C ~
                         ChemPhys_lake_subset$lakename)
lake_twosample

# format as a GLM
lake_twosample2 <- lm(ChemPhys_lake_subset$temperature_C ~
                      ChemPhys_lake_subset$lakename)
lake_twosample2
```

>Answer: 
